---
title: "ECO 395M Final Project"
author: "Hannah Jones & Joey Herrera"
date: "5/10/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Abstract


Improving student success via better grades is a focus of every school because better grades give student a higher chance at having a better quality of life. One factor associated with good grades is high parent and family involvement in a students academic journey. Thus, many schools emphasize increasing parent and family involvement to facilitate better student outcomes. This paper seeks to provide schools with equitable program recommendations using data from the NHES's parent and family involvement survey. This data will be evaluated using principal component analysis and fitting a random forest model to extract meaningful summaries of relevant features.

- Section on findings

Questions trying to answer: 
- How does family engagement/parent engagement impact student outcomes?

## Introduction

Introduce data, introduce problem, why it is important


Paragraph draft: Student success, which is measured through standardized test scores and grades, is the universal measuring stick for enhanced opportunity in the United States. One of the most significant indicators in predicting student success is parent and family involvement in a student's life. Since schools only facilitate learning to students throughout the day, a burden is placed on parents and legal guardians to continue cultivating a learning environment outside of school. To ease the burden and increase family engagement, schools have created programs and events, such as school carnivals, dances, and bring your parent to class day. These traditional programs and events seem to work well in improving the family involvement for students whose parents work traditional 9-5 jobs. Most families of color and low-income households are not afforded this same opportunity,

Creating an equitable array of programs to increase parent and family involvement for families of different races, ethnicities, and socioeconomic statuses is a growing issue. In response, the National Household Education Surveys (NHES) program created a national survey with several hundred questions focusing on different aspects of parent and family engagement, This survey is sent to households across the United States, where parents have the opportunity to fill it out and return their responses. The NHES typically receives around 50 million survey responses each time the survey is sent out. 

This paper seeks to provide basic program recommendations to schools by evaluating the impact of indicators derived from the NHES parent and family involvement (ppfi) survey on student grades. Using a 1% sample of the full data set, this paper will use a principal component analysis (PCA) and random forest forest technique to illustrate the effects of different pfi question (or features) on student grades.  We hope to see which demographic characteristics are highly predictive in success or lack of success, and how this information can be used to target students at risk of falling behind.

----
- find lit/article about what has been done to bolster parent enagagement, and how lack of parent engagement affects children generally, not necesarily just through school ---> Visioning Parent Engagement in Urban Schools, https://journals.sagepub.com/doi/pdf/10.1177/105268460701700602
      - Parents, in particular, parents of color support their children in ways that go undetected by schools.
      - Parents—especially, low-income and minority parents—are more likely to be involved in education when schools invite their participation, provide multiple entry points for involvement, value their perspectives, and reach out in culturally appropriate ways.
      - Researchers have noted the ways in which schools marginalize parents of color and low-income parents, and they have pointed to the limitations of traditional conceptions of parent involvement, geared to the school’s agenda.
      
Paragraph draft: Given the importance of family involvement in student success, schools have put resources into creating programs that address the specific needs based on the demographic and socioeconomic information of student families. In particular, low-income and minority parents are less likely to be involved with their students' education in a traditional sense. Various structural factors including working multiple jobs or obstructive job hours. To provide these demmographics with alternate opportunities schools have provided  multiple opportunities for invovement, value different perspectives, and reach out to families in culturally appropriate ways (Auerbach, 2007). Schools must continue to adapt to their individual demographics because traditional concepts of family involvment limit the opportunity for low-income and minority-families to be involved. Thus, those students are likely to have less academic success than their counterparts.
---- 

- find lit/article about growing space for ML in education ---> The Importance of Machine Learning in the Education Sector, https://www.newgenapps.com/blog/the-importance-of-machine-learning-in-the-education-sector/#:~:text=Improving%20the%20Efficiency%20of%20Both,manage%20their%20content%20and%20curriculum.&text=Also%2C%20Machine%20Learning%20makes%20educators,classroom%20management%2C%20scheduling%2C%20etc.

Paragraph draft: An efficient way to provide recommendations to schools and policymakers is through the involvement of machine learning techniques. These data science skills have the power to delineate trends from large data sets, such as data revolving around the number of students in a school, in an interpretable manner. Besides predicting student success, there are a variety of places that machine learning can enhance the education sector. For instance, creating adaptive learning for people with different learning styles, improving efficiency in the learning environment, and assessment evaluation are just a few ways machine learning can postivly impact the education sector (Ryan, 2020).

----
- discussion of how using creative ML/data mining can lead to more robust/actionable/empathetic understandings of students, beyond simply looking at grades and attendance-- contextual understanding of kids/kids success/family opportunities to elicit success


----
Transition into the rest of the paper

The following sections of this paper will address the empirical methodology, results, and concluding remarks associated with the research problem. 

=======
- find lit/article about what has been done to bolster parent enagagement, and how lack of parent engagement affects children generally, not necesarily just through school
- find lit/article about growing space for ML in education
- discussion of how using creative ML/data mining can lead to more robust/actionable/empathetic understandings of students, beyond simply looking at grades and attendance-- contextual understanding of kids/kids success/family opportunities to elicit success


## Methods

Use Principle Components Analysis to see how different traits/features group.  Allows us to recommend a collection of actions versus more simple recommendations.



- PCA 
- Random forest

(look at clustering too)

## Results 

Interesting things to look at:
- Individual features (holding all else constant)
- PCA-- how do things group/cluster

Show:


```{r}


```


## Conclusion


## Appendix


## Bibliography

  Auerbach, Susan. “Visioning Parent Engagement in Urban Schools.” Journal of School Leadership, vol. 17, no. 6, Nov. 2007, pp. 699–734, doi:10.1177/105268460701700602.

Ryan, Jacob. “The Importance of Machine Learning in the Education Sector.” NewGenApps, 20 Dec. 2020, www.newgenapps.com/blog/the-importance-of-machine-learning-in-the-education-sector/#:~:text=Improving%20the%20Efficiency%20of%20Both,manage%20their%20content%20and%20curriculum.&amp;text=Also%2C%20Machine%20Learning%20makes%20educators,classroom%20management%2C%20scheduling%2C%20etc. 

