---
title: "ECO 395M Final Project"
author: "Hannah Jones & Joey Herrera"
date: "5/10/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Abstract

Questions trying to answer: 
- How does family engagement/parent engagement impact student outcomes?

## Introduction

Introduce data, introduce problem, why it is important
- find lit/article about what has been done to bolster parent enagagement, and how lack of parent engagement affects children generally, not necesarily just through school
- find lit/article about growing space for ML in education
- discussion of how using creative ML/data mining can lead to more robust/actionable/empathetic understandings of students, beyond simply looking at grades and attendance-- contextual understanding of kids/kids success/family opportunities to elicit success

## Methods

Use Principle Components Analysis to see how different traits/features group.  Allows us to recommend a collection of actions versus more simple recommendations.



- PCA 
- Random forest

(look at clustering too)

## Results 

Interesting things to look at:
- Individual features (holding all else constant)
- PCA-- how do things group/cluster

Show:


```{r}


```


## Conclusion


## Appendix

